Fail Faster

### 2025-03-05 paper/paper.tex
things
skeleton further

\section{Background}
\subsection{Staged Metaprogramming}

@@

\subsection{Property-Based Testing}
\subsection{Other Languages and Libraries}

@@

\jwc{
  \begin{itemize}
    \item How does BQ work?
    \begin{itemize}
      \item The basic generator type: \texttt{'a generator = size:int -> random:SR.t -> 'a}. Size and random seed to deterministic value.
      \item Also two inverse functions, \texttt{create}, and \texttt{generate}.
      \item This gets a monad intance in the obvious way (show code).
    \end{itemize}
    \item Note that \emph{extensionally} \texttt{generate (create (fun ~size ~random -> e)) ~size ~random = e}, but the OCaml compiler does not always perform this optimization, or do the inlining required to expose it.
    (When sufficiently obfuscated behind returns and binds ...) This program compiles to code that (1) allocates the closure for `e', (2) passes it to create (which returns the closure), and then calls (3) generate, which immediately jumps into the closure.
  \end{itemize}
}

In OCaml, garbage collection is the killer.

This generator generates a list of bools. To generate a list of 100 bools, it allocates \texttt{4808.00w (minor)}, and runs for an average of
3.9us. Merely representing a list of bools of length 100 requires 300 words in OCaml, so this generator allocates 16 times more than the theoretical minimum.

\begin{lstlisting}
let gen_list_bool =
  fixed_point (fun gl ->
    let%bind n = size in
    if n <= 0 then return [] else
      let%bind x = bool in
      let%bind xs = with_size ~size:(n-1) gl in
      return (x :: xs)
  )
\end{lstlisting}

On the other hand, this generator allocates only 303w, and runs for 0.7us, to generate the same list of 100 bools.
\begin{lstlisting}
  let rec gen_list_bool_faster ~size ~random  = 
    if size <= 0 then []
    else (Unboxed_splitmix.bool random)::(gen_list_bool_faster ~size:(size - 1) ~random)
\end{lstlisting}

@@

\subsubsection{Expensive Combinators}
\jwc{Many combinators like \texttt{union} and \texttt{weighted_union} needlessly allocate large data structures in the generation hot path.
Using a combination of staging and more careful algorithm design, we can avoid this overhead.
}
% \subsubsection{Function call overhead}
\subsubsection{Boxed integers}

@@

\jwc{Here!}

\begin{lstlisting}
type 'a gen = int code -> Random.t code -> 'a
\end{lstlisting}

\tr{Now, we formally introduce the library, describe all its functionality, how things work, etc.}

>>>>

\section{Background}

\subsection{Property-Based Testing}
\jwc{Mention the shape of PBT dsls, give running example here (or maybe earlier!). Monadic dsls.}

\begin{lstlisting}
let int_pair : (int * int) Bq.t =
  let%bind x = int 0 100 in
  let%bind y = int 0 x in
  return (x,y)
\end{lstlisting}

\jwc{\texttt{int} here is really \texttt{int\_uniform\_inclusive}. I also think we should completely omit named arguments in the paper so we don't confuse with splices.}

\subsection{Staged Metaprogramming}

@@

\subsection{Other Languages and Libraries}

@@

\jwc{
  \begin{itemize}
    \item How does BQ work?
    \begin{itemize}
      \item The basic generator type: \texttt{'a generator = int -> SR.t -> 'a}. Size and random seed to deterministic value. (note: \texttt{SR.t} is a mutable seed)
      \item Also two inverse functions, \texttt{create}, and \texttt{generate}.
      \item This gets a monad intance in the obvious way (show code).
      \item Also show the code for \texttt{int}, how it calls the underlying SR function.
    \end{itemize}
    \item Note that \emph{extensionally} \texttt{generate (create (fun size random -> e)) size random = e}, but the OCaml compiler does not always perform this optimization, or do the inlining required to expose it.
    (When sufficiently obfuscated behind returns and binds ...) This program compiles to code that (1) allocates the closure for `e', (2) passes it to create (which returns the closure), and then calls (3) generate, which immediately jumps into the closure.
    \item Show allocation benchmarks of the running example, versus the version where you inline everything.
  \end{itemize}
}

% In OCaml, garbage collection is the killer.

% This generator generates a list of bools. To generate a list of 100 bools, it allocates \texttt{4808.00w (minor)}, and runs for an average of
% 3.9us. Merely representing a list of bools of length 100 requires 300 words in OCaml, so this generator allocates 16 times more than the theoretical minimum.

% \jwc{Mention ppx_jane and let syntax}

% \begin{lstlisting}
% let gen_list_bool =
%   fixed_point (fun gl ->
%     let%bind n = size in
%     if n <= 0 then return [] else
%       let%bind x = bool in
%       let%bind xs = with_size ~size:(n-1) gl in
%       return (x :: xs)
%   )
% \end{lstlisting}

% On the other hand, this generator allocates only 303w, and runs for 0.7us, to generate the same list of 100 bools.
% \begin{lstlisting}
%   let rec gen_list_bool_faster ~size ~random  = 
%     if size <= 0 then []
%     else (Unboxed_splitmix.bool random)::(gen_list_bool_faster ~size:(size - 1) ~random)
% \end{lstlisting}

@@

\subsubsection{Expensive Combinators}
\jwc{Many combinators like \texttt{union} and \texttt{weighted\_union} needlessly allocate large data structures in the generation hot path.
Using a combination of staging and more careful algorithm design, we can avoid this overhead.
}

\begin{lstlisting}
  let of_weighted_list alist =
  let weights, values = List.unzip alist in
  let value_array = Array.of_list values in
  let total_weight, cumulative_weight_array =
    let array = Array.init (Array.length value_array) ~f:(fun _ -> 0.) in
    let sum =
      List.foldi weights ~init:0. ~f:(fun index acc weight ->
        let acc' = acc +. weight in
        array.(index) <- acc';
        acc'
      )
    in
    sum, array
  in
  create (fun ~size:_ ~random ->
    let choice = Splittable_random.float random ~lo:0. ~hi:total_weight in
    match
      Array.binary_search
        cumulative_weight_array
        ~compare:Float.compare
        `First_greater_than_or_equal_to
        choice
    with
    | Some index -> value_array.(index)
    | None -> assert false
   )
\end{lstlisting}

\jwc{This builds the cdf of the distribution, samples from 0 to the total, and binary searches through the array to find the bucket.}
\jwc{Allocating this array is unnecessary: we can simply compute the total weight, sample \texttt{x} between 0 and the total, and then walk the list accumulating the sum, until the accumulator exceeds \texttt{x}. Since in practice the list of possible options is quite small (usually at most 10 in practice), the linear time scan is going to be much faster.}
\jwc{(The point of this para will just be to emphasize that it's almost always better to be allocation-aware, instead of algorithmically clever.)}

\jwc{Moreover, the possible options is almost always statically known in practice: you use weightedunion to generate (say) a datatype with possible
variants, whose options are known at compile time.
Put more simply, you basically always call weighted union with an \emph{explicit list}: For this reason, you should not have to incur the cost of allocating this list at runtime.}

% \subsubsection{Function call overhead}
\subsubsection{Boxed integers}

@@

\subsection{Basic Design}

\begin{lstlisting}
type 'a gen = int code -> Random.t code -> 'a code

let return (cx : 'a code) : 'a gen = fun size_c random_c -> cx

let bind (g : 'a gen) (k : 'a -> 'b gen) : 'b gen =
  fun size_c random_c ->
    .<
      let x = .~(g size_c random_c) in
      .~(
        (k .<x>.) size_c random_c
      )
    >.

let int (lo : int code) (hi : int code) : int code gen =
  fun size_c random_c ->
    .< SR.int .~random_c .~lo .~hi >.
\end{lstlisting}

\jwc{Conversion to a bq generator}

\begin{lstlisting}
let to_bq (g : 'a code gen) : ('a Bq.t) code =
  .<
    Bq.create (fun size random ->
      .~(g .<size>. .<random>.)
    )
  >.
\end{lstlisting}

\jwc{Then show the running example, with }

\begin{lstlisting}
let int_pair_staged : ((int * int) code) gen =
  let%bind x = int 0 100 in
  let%bind y = int 0 x in
  return (x,y)

let int_pair : (int * int) Bq.t code = to_bq int_pair_staged
====
.<
  Bq.create (fun size random ->
    let x = SR.int random 0 100 in
    let y = SR.int random 0 x in
    (x,y)
  )
.>
\end{lstlisting}

\jwc{So if you write with the \texttt{gen} combinators and then only call \texttt{to\_bq} at the end, you get something that includes no creates/generates!}

\jwc{Show benchmark difference between these two generators.}

\subsection{Staging More Combinators}

\jwc{Weighted unions: recall that we should not need to materialize}

\subsection{Basic Design}
